A radiocarbon data based evaluation of Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe

### 2018-11-27 01_Introduction.Rmd
spelling

This articles employs some of the methods that arose within archaeology's contemplation of *Cultural transmission theory* to trace the development of four major bronze age burial traditions in Europe. The cultural-historical evolution of archaeological cultures and ethnic groups during the european bronze age is complex and it's reconstruction an ongoing endeavour of archaeology. Within this history of constant emergence, displacement and demise many different forms of burial traditions were practiced. Most of them can be distinguished by the criteria *burial construction* and *treatment of the body*. Bronze age burials are either flat graves or part of a burial mound, and they contain either untreated or cremated human remains. Openly available bulk radiocarbon data now allows to reconstruct the trends within the distribution of these complex behaviours to test classical archaeological narratives about their respective prevalence. The comparison of macroregional trends also opens the perspective on long term patterns of interaction intensity on a continental scale.

@@

Archaeology has a special perspective to offer for Cultural evolution research: It's concerned with questions that can embrace continent size areas and millennia of cultural development [@garveyCurrentPotentialRoles2018]. Archaeological data is mostly comprised of the material remains of human behaviour -- so the ruins of past architecture, individual lost or discarded artefacts and all more inconspicuous by-products of living. We rightfully expect humans to learn handcraft as well as building techniques and traditions  -- both functionally meaningful and seemingly pointless -- from others close to them. The similarity of historically and spatially close material remains can therefore be explained as a result of social interaction [@lymanChronometersUnitsEarly2000; @lymanCultureHistoricalBiological2001; @lymanRiseFallCulture1997; @obrienEpistemologicalNatureArchaeological2002]. In this context a distinction of culture in function and style became an important question for *Evolutionary archaeology*, explicitly elaborated already by @dunnell1978style. Some types of human behaviour obviously increase wellbeing and simplify survival, like certain hunting techniques, superior tools or sowing more fruitful crops. Without resorting to the inappropriate simplifications of Evolutionism their diffusion can be expected to increase over time. On the other hand there are behaviours that are not obviously related to survival and reproductive capacity, like e.g. ceramic decorations, traditional costume fashion or jewelry design. Despite their seemingly secondary importance these characteristics often pass through a breathtakingly fast development and serve archaeologists as proxy for relative chronology. Dunnell postulates that the distribution and emergence of these stylistic cultural traits behave randomly on a population scale level. The stochastic processes that are at work here are an important object of investigation and discussed as *Neutral Transmission* or *Stylistic variability* [@bentleyRandomDriftCulture2004; @lipoNeutralityStyleDrift2001; @neimanStylisticVariationEvolutionary1995; @rogersNaturalSelectionCultural2008; @shennanCeramicStyleChange2001].

@@

Evolutionary Archaeology has - parallel to the further development of the the science of archaeology - in the past 15 years of dynamism. The latest Developments show that they are more successful than almost any other archaeological faculty the scientific and Ideal for gradual, theoretical further development from one contribution to the next. During Neiman the concept of his simulation from population genetics was borrowed from other methodology for the Preparation and investigation of phylogenetic trees from cladistics to archaeological questions applied . Stylistic Variability is a very promising basis for modelling with regard to cultural (in comparison to spatial) distance. The Bayes' theorem has proven to be a potent statistical tool. to the model analysis and evaluation revealed . Archaeological data sets and their special features for Cultural evolution research is becoming more rigorous today. questioned and analyzed more consciously - beyond the anecdotal evidence.

>>>>

This articles employs some of the methods that arose within archaeology's contemplation of *Cultural transmission theory* to trace the development of four major bronze age burial traditions in Europe. The cultural-historical evolution of archaeological cultures and ethnic groups during the European bronze age is complex and it's reconstruction an ongoing endeavour of archaeology. Within this history of constant emergence, displacement and demise many different forms of burial traditions were practised. Most of them can be distinguished by the criteria *burial construction* and *treatment of the body*. Bronze age burials are either flat graves or part of a burial mound, and they contain either untreated or cremated human remains. Openly available bulk radiocarbon data now allows to reconstruct the trends within the distribution of these complex behaviours to test classical archaeological narratives about their respective prevalence. The comparison of macroregional trends also opens the perspective on long term patterns of interaction intensity on a continental scale.

@@

Archaeology has a special perspective to offer for Cultural evolution research: It's concerned with questions that can embrace continent size areas and millennia of cultural development [@garveyCurrentPotentialRoles2018]. Archaeological data is mostly comprised of the material remains of human behaviour -- so the ruins of past architecture, individual lost or discarded artefacts and all more inconspicuous by-products of living. We rightfully expect humans to learn handcraft as well as building techniques and traditions  -- both functionally meaningful and seemingly pointless -- from others close to them. The similarity of historically and spatially close material remains can therefore be explained as a result of social interaction [@lymanChronometersUnitsEarly2000; @lymanCultureHistoricalBiological2001; @lymanRiseFallCulture1997; @obrienEpistemologicalNatureArchaeological2002]. In this context a distinction of culture in function and style became an important question for *Evolutionary archaeology*, explicitly elaborated already by @dunnell1978style. Some types of human behaviour obviously increase well-being and simplify survival, like certain hunting techniques, superior tools or sowing more fruitful crops. Without resorting to the inappropriate simplifications of Evolutionism their diffusion can be expected to increase over time. On the other hand there are behaviours that are not obviously related to survival and reproductive capacity, like e.g. ceramic decorations, traditional costume fashion or jewellery design. Despite their seemingly secondary importance these characteristics often pass through a breathtakingly fast development and serve archaeologists as proxy for relative chronology. Dunnell postulates that the distribution and emergence of these stylistic cultural traits behave randomly on a population scale level. The stochastic processes that are at work here are an important object of investigation and discussed as *Neutral Transmission* or *Stylistic variability* [@bentleyRandomDriftCulture2004; @lipoNeutralityStyleDrift2001; @neimanStylisticVariationEvolutionary1995; @rogersNaturalSelectionCultural2008; @shennanCeramicStyleChange2001].

@@

Evolutionary Archaeology has - parallel to the further development of the the science of archaeology - in the past 15 years of dynamism. The latest Developments show that they are more successful than almost any other archaeological faculty the scientific and Ideal for gradual, theoretical further development from one contribution to the next. During Neiman the concept of his simulation from population genetics was borrowed from other methodology for the Preparation and investigation of phylogenetic trees from cladistics to archaeological questions applied . Stylistic Variability is a very promising basis for modelling with regard to cultural (in comparison to spatial) distance. The Bayes' theorem has proven to be a potent statistical tool. to the model analysis and evaluation revealed . Archaeological data sets and their special features for Cultural evolution research is becoming more rigorous today. questioned and analysed more consciously - beyond the anecdotal evidence.
